End-User Software Customization by Direct Manipulation of Tabular Data

### 2020-05-23 paper.tex
first cut of final draft
possible final

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{End-User Customization by Direct Manipulation of Tabular Data}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  In this paper we introduce \emph{table-driven customization}, a new
  way for end users to customize software. We augment a user interface
  with a table view showing the structured data inside the application.
  When users edit the table, their changes are reflected in the original
  UI. This simple model also accommodates a spreadsheet formula language
  and custom data editing widgets, providing enough power to implement a
  variety of useful extensions.

@@

Many end user customization systems
\citep{cook2007, bolin2005, leshed2008, chasins2018} offer a scripting
model. They use various strategies to make programming more
approachable: friendly syntax, a visual programming environment, or
macro recording to bootstrap from concrete demonstrations. But all these
techniques build on the same fundamental foundation: an imperative
programming model, with statement sequencing, mutable variables, and
loops.

We have known for decades about an alternative: \emph{direct
manipulation} \citep{shneiderman1983}, where ``visibility of the object
of interest'' replaces ``complex command language syntax''. Direct
manipulation is the \emph{de facto} standard in GUIs today, but when it
comes to customizing those GUIs, it is rarely to be found. Switching
from using an interface to customizing it via scripting requires an
abrupt shift in interaction model, and can pose a steep learning barrier
for users not familiar with programming. We subscribe to MacLean et al's
vision of a ``gentle slope'' \citep{maclean1990}, where users should
only need to make minimal and incremental investments in skill to
achieve their desired customizations.

We seek to add a point to this gentle slope by proposing a new method
for customizing software via direct manipulation, which requires less
skill than scripting but still supports powerful customizations. We take
inspiration from visual database query interfaces and spreadsheets,
which have successfully enabled millions of end users to compute with
data through direct manipulation.

In this paper, we present a paradigm called \emph{table-driven
customization}. An application's UI is augmented with a table view where
the user can see and manipulate the application's internal data, just as
in a familiar spreadsheet. These changes don't just apply to the table;
they also result in immediate changes to the application's original user
interface. The user can sort/filter data in the UI, inject annotations
into the UI, pull in related information from other web services, and
more, all using the table as a mediating interface.

@@

We have successfully used Wildcard to build customizations for 11
different websites which serve our own personal needs. In
Section~\ref{sec:reflections}, we present reflections from this process.
We outline the kinds of customizations we were able to build,
limitations we encountered, and some of the challenges of writing
scraping logic.

@@

\begin{itemize}
\tightlist
\item
  \emph{Customization by direct manipulation}: We agree with MacLean et
  al \citep{maclean1990} that customization should follow a ``gentle
  slope'', where users can gradually invest in more complex techniques
  to achieve deeper customizations. We believe an important point on
  this slope is the ability to customize an application by directly
  seeing and changing its internal structured data, rather than by
  writing imperative scripts.
\item
  \emph{Wrapping applications for customization}: Typically, tools that
  don't rely on official extension APIs resort to offering low-level
  APIs for customization. Instead, we propose a community-maintained
  library of semantic wrappers around existing applications, enabling
  end users to work with domain objects rather than low-level
  representations.
\end{itemize}

Table-driven customization relates to existing work in many areas. In
particular, our goals overlap with many software customization tools,
and our methods overlap with direct manipulation interfaces for working
with structured data, including visual database query systems and
spreadsheets. We explore these connections and more in
Section~\ref{sec:related-work}.

@@

To illustrate the end user experience of table-driven customization, we
consider an example scenario of customizing
\href{https://news.ycombinator.com/}{Hacker News}, a tech news
aggregator. Figure~\ref{fig:hacker-news} shows accompanying screenshots.

@@

To achieve this ordering, the user simply clicks on the ``points''
column header in the table. This sorts the table view by points, and the
website UI also becomes sorted in the same order
(Figure~\ref{fig:hacker-news}, Note B)---Wildcard has manipulated the
website's DOM to synchronize it with the sort order of the table. This
sort order is also persisted in the browser and reapplied automatically
the next time the user loads the page, so they can always browse the
latest set of links using this sort order.

\textbf{Adding estimated read times}: Next, the user decides to attempt
a more substantial customization: to add estimated read times to
articles on Hacker News.

The table contains additional empty columns (\texttt{user1},
\texttt{user2}, \ldots), where the user can enter spreadsheet-like
formulas to compute derived values. The user enters a formula into the
first column: (Figure~\ref{fig:hacker-news}, Note C):

@@

This calls a built-in function that takes a URL as an argument and uses
a public web API to compute an estimated read time for the link's
contents. The \texttt{link} argument refers to a column name in the
table; the formula is automatically evaluated across all rows in the
table, using the value of \texttt{link} for each row.

The user clicks the \texttt{user1} column header to sort the articles on
the page in descending order of estimated read time, helping to
prioritize reading deeper content. It would be useful to show these read
times in the page too, but before doing that, the user wants to make the
time estimates more readable. They enter another formula in the next
column, \texttt{user2}:

@@

This formula converts seconds to minutes by dividing by 60 and rounding
to the nearest integer, and then concatenates a label to the number,
producing strings like ``21 min read''.

Finally, the user clicks a menu option in the table header to display
the contents of this new column in the original page
(Figure~\ref{fig:hacker-news}, Note D). Each article on the page now
shows an annotation with the estimated read time in minutes. (The format
of how annotations appear for each item was determined by the programmer
who created the Wildcard adapter for Hacker News.)

\textbf{Adding notes to links}: The user can also manually add notes to
the table, by simply entering values into the table without using
formulas. In this case, the user jots down a few notes in another column
about articles they might want to read, and the notes appear in the page
next to the read times (Figure~\ref{fig:hacker-news}, Note D). The
annotations are also stored in the browser's local storage so they can
be retrieved on future visits.

@@

They can then filter the table to only contain rows where this formula
column contains \texttt{false}; visited rows are hidden both from the
table view and the original page. This is an example of a customization
that the original website could not implement, since websites don't have
access to the browser history for privacy reasons. But by using
Wildcard, the user can implement the customization locally, without
needing to expose their entire browser history to Hacker News.

@@

Figure~\ref{fig:table-adapter} summarizes the overall architecture of
table-driven customization, using a simplified illustration of the
Hacker News example scenario. In this example, the name and points value
for each article is scraped from the web page DOM, and user annotations
are loaded from the brower's local storage.

First, the web page and the browser storage are each wrapped by a
\textbf{table adapter}, which defines a bidirectional mapping between
some underlying data source and a table.

In addition to a \emph{read} mapping for how the underlying data should
be represented as a table, it also defines a \emph{write} mapping for
what effects edits should have on the original data source. In this
case, the local storage adapter has a trivial mapping: it loads a table
of data stored in the browser, and persists edits to that persisted
state. The mapping logic of the DOM scraping adapter is much more
involved. It implements web scraping logic to produce a table of data
from the web page, and turns edits into DOM manipulations, such as
reordering rows of data on the page.

@@

The first two parts of the table adapter interface resemble a typical
database table:

@@

\emph{Handling edits}: The query engine can issue a request to a table
adapter to make an edit to a record. The meaning of making an edit can
vary depending on the adapter: in the local storage adapter, an
annotation may be persisted into local storage; in the DOM scraping
adapter, an edit may represent filling in a form field.

@@

\emph{Data from other tables}: The query engine provides each table
adapter with the entire combined table shown to the user. The DOM
scraping adapter uses this for injecting annotations---values in
additional columns from other tables are added to the original web page.

\emph{Currently selected record}: As the user clicks in different parts
of the table view, the query engine communicates the record currently
selected by the user. The DOM scraping adapter uses this information to
highlight the row in the page that corresponds to the selected row in
the table (and scroll to the row if it's not currently showing on the
page), which helps clarify the connection between the table and the
original UI.

@@

Because each website has unique content, we rely on programmers to
create a DOM scraping adapter for each individual website to make it
available for customization in Wildcard. To make this approach viable,
we have built a generic DOM scraping adapter which a programmer can
configure with the minimal site-specific parts.

\emph{Specifying scraping logic}: First, the adapter author specifies
basic metadata like which URLs the adapter should apply to, and a list
of columns and their types. Then, implement a single Javascript function
that extracts relevant elements from the page using web scraping
technqiues. The generic adapter then wraps this scraping function to
implement the table adapter interface. For example, when the user sorts
the table view, the generic adapter removes the DOM elements
corresponding to the rows in the table, and reinserts them back into the
page in the correct order.

For each scraped row of data, the programmer must return a unique
identifier, and a value for each column in the table.

\emph{Choosing an ID}: When possible, we recommend choosing a
server-side identifier that will remain stable across pageloads. This
enables user annotations persisted in local storage to be associated
with the same records on subsequent pageloads. We have found that it's
usually possible to find such an identifier; for example, each item in a
page often contains a link to a page with more details, where the URL
contains a unique identifier.

@@

The default option is to return a DOM element, in which case the generic
adapter extracts the text contents of the DOM element and casts them to
the type of the column. For example, if the relevant column is a numeric
column and the returned value is a \texttt{\textless{}div\textgreater{}}
containing the string ``1.23'', the generic adapter extracts the number
1.23. The advantage of this approach is that the field becomes
\emph{writable} in the table, if the adapter author chooses to mark it
as writable. When the user edits the value in the table, the generic
adapter can overwrite the inner contents of the DOM element to reflect
the change. In practice, this is most useful for editing the value of
form input fields. It's possible to enable the editing of other fields
(e.g., editing the title of a post on Hacker News) but we haven't found
many use cases for editing those types of fields.

Another option is to directly return a value, rather than returning a
DOM element. The advantage of this approach is that the adapter author
can perform arbitrary computations to derive the returned value---for
example, they can use a regular expression to extract a substring. The
disadvantage is that the field is no longer writable, because the
computation used to derive the value isn't reversible.

@@

As with DOM scraping adapters, we have made it easy for programmers to
create site-specific AJAX scraping adapters. A programmer writes a
function that specifies how to extract data from an AJAX request, and
the framework handles the details of intercepting requests and calling
the programmer-defined function.\footnote{So far we have only
  implemented AJAX scraping in the Firefox version of Wildcard, since
  Firefox has convenient APIs for intercepting requests. It appears
  possible to implement in Chrome as well, but we have not finished our
  implementation.}

@@

With the advent of rich frontend web frameworks, structured application
state is now often available in the web client. We suspect it is
possible to create plugins for frontend frameworks that expose this
state to Wildcard with only minimal effort from the application
developers. To test this hypothesis, we have created an early prototype
of a Wildcard adapter for the Redux state management library, and used
it with the TodoMVC todo list application. To install the adapter, the
programmer only needs to make a few small changes to their app:

\begin{itemize}
\tightlist
\item
  \emph{Exposing a table}: In Redux, app developers already represent
  the state of a user interface as a single object. To expose a table of
  data, the developer simply writes a function that maps the state
  object to a table.
\item
  \emph{Editing data values}: The developer defines how edits to records
  in the table should map to edits in the state object.
\item
  \emph{Sorting/filtering the UI}: The developer inserts an additional
  data processing step before the state object is rendered in the UI, so
  that Wildcard can apply sorting and filtering.
\item
  \emph{Annotations}: The developer adds additional React components
  managed by Wildcard into their UI tree, which render additional data
  as annotations.
\end{itemize}

@@

One way to view this query model is as a tiny subset of the SQL query
model. Despite its simplicity, this model has proven sufficient for
meeting the needs of customization in practice, and minimizes the
complexity of supporting arbitrary queries. But because it fits into the
general paradigm of relational queries, it could theoretically be
extended to support a wider range of queries.

The query engine is also responsible for executing formulas. We have
built a small formula language resembling a spreadsheet formula
language, except that formulas automatically apply across an entire
column of data, and reference other column names instead of values in
specific rows. This is more convenient than needing to copy-paste a
formula across an entire column as in spreadsheets, and has worked for
all of the customizations we have built.

@@

The table editor only serves as a shallow interface layer over the query
engine, relaying user commands to the query engine and rendering the
resulting data table. Because of this architectural split, it would be
straightforward to develop additional table editor interfaces on top of
the Wildcard system. One intruiging future possibility is to apply the
idea of cell editors to the entire table, by developing table editors
which don't look like a table. For example, we could provide a calendar
view for displaying a table containing a date column.

@@

We built Wildcard integrations for 11 websites, including transactional
sites like Amazon and Uber Eats, and media consumption sites like
Youtube and Hacker News. Table 1 summarizes these integrations, showing
the number of lines of code in the adapter configuration for each site,
and some example customizations we were able to perform for each site.

So far, most usage of Wildcard has come from members of the project
team. Here we offer our reflections on using the system, focused on two
key questions:

@@

Sometimes, websites have opaque ranking algorithms which presumably
maximize profit but don't offer much control to the user. For example,
Airbnb previously allowed users to sort listings by price, but removed
that feature in 2012. We have found other sites where a lack of sorting
options seems more like merely an omission; for example, the Instacart
grocery delivery service has a spartan UI for viewing a pending order,
which doesn't allow for sorting by price or category. In both of these
cases, Wildcard enables users to take back some control.

In the current implementation of Wildcard, users can only sort and
filter entries that are shown on the current page, which means that
users are not entirely liberated from the suggestions of the opaque
algorithm. This restriction could be overcome in the future by scraping
content across multiple pages, or by using an integrated adapter built
in to the application. However, we've also found that sorting/filtering
a single page of a paginated list is often an acceptable outcome (and
sometimes even a preferable one). It's more useful, for example, to sort
30 recommended Youtube videos than to try to sort all videos on Youtube.

\begin{figure}
\hypertarget{fig:amazon}{%
\centering
\includegraphics[width=\columnwidth]{media/amazon.png}
\caption{Sorting the used sellers page on Amazon by total price}\label{fig:amazon}
}
\end{figure}

\begin{figure}
\hypertarget{fig:ubereats}{%
\centering
\includegraphics[width=\columnwidth]{media/ubereats.png}
\caption{Organizing takeout restaurants on Uber Eats by ETA and price}\label{fig:ubereats}
}
\end{figure}

@@

Many web annotation systems focus on annotating text or arbitrary
webpage content, but Wildcard limits annotations to structured objects
extracted by an adapter, resulting in a different set of use cases.
Annotating with Wildcard has proven most useful when taking notes on a
list of possible options (e.g., evaluating possible Airbnb locations to
rent). We have also used it with Instacart's online grocery cart, for
jotting down notes like ``do we need more eggs?''

@@

We developed two cell editors to explore different benefits of adding
custom UI for editing the table.

An important benefit of cell editors is that they allow users to
incorporate their private information into a web UI without uploading it
to a website. To explore this idea, we created a datepicker based on the
\href{https://fullcalendar.io/}{FullCalendar} plugin, which can load
data from a Google Calendar. This makes it convenient to enter dates
into a website based on the user's personal calendar information.

Another benefit of cell editors is that a user can choose their
preferred widget for editing some type of information. We built a text
editor based on the \href{https://ckeditor.com/}{CKEditor} rich text
editor, and demonstrated its use by integrating it with Google's Blogger
website and using it to edit blog post content.

@@

AJAX scraping proved highly useful in several cases. The Uber Eats
website was challenging to scrape because it has a complex DOM structure
with machine-generated CSS classes, but the site also use AJAX requests
which contain all the relevant data in a structured form that is much
easier to extract. We also found examples where relevant information
wasn't present in the DOM at all. On the grocery delivery site
Instacart, we found that AJAX requests contained additional data not
shown in the UI, enabling us to do things like sort grocery items in an
order by category.

@@

Here we discuss some themes that we've explored with this work, which go
beyond our specific approach and address issues in software
customization more broadly.

@@

With table-driven customization we aim to provide a gentler slope, by
using direct manipulation for software customization. The data shown in
the table view is the domain data from the original application. The
user makes changes to the data by selecting areas of interest in the
table, e.g.~sorting/filtering by clicking the relevant column header, or
adding annotations by clicking on the relevant row. These interactions
are common in GUI applications, and Wildcard therefore meets MacLean et
al's goal: some one-click customizations are as easy as using the
original application. Formulas introduce some additional complexity, but
the pure functional semantics and the intermediate visibility of results
make the task still easier than imperative programming.

One aspect of directness that we have chosen not to maintain in Wildcard
is enabling customization in closer proximity to the original user
interface elements, as explored by tools like Scotty \citep{eagan2011}.
We have found that augmenting the original UI with a distinct,
additional representation provides a more consistent experience across
all applications, and clearly shows the user what structured data is
available to work with.

Ainsworth et al provide a helpful taxonomy of the value of multiple
representations \citep{ainsworth1999}. In their terms, Wildcard plays a
\emph{complementary role} by supporting a \emph{different set of tasks}
from the original application, while displaying \emph{shared
information}. Wildcard may also help construct \emph{deeper
understanding by subtraction}. By stripping away details and only
showing the essential data in an interface, Wildcard helps users think
of new ways of using that data, outside the specific constraints of the
original application.

As noted in Section~\ref{sec:reflections}, there are many customizations
that can be achieved in scripting languages that cannot be implemented
in Wildcard. We consider this an acceptable tradeoff in exchange for a
gentler slope in customization, and we show in Section~\ref{sec:example}
and Section~\ref{sec:reflections} that our model can still implement
many useful customizations.

@@

\emph{Anticipated customization tools}, in contrast, use explicit
extension APIs provided by the application developer. Examples of this
include accessing a backend web API, or writing a customization in
Applescript for an application that exposes its domain model to the
scripting language. The main benefit of this style is that it allows the
extension author to work with meaningful concepts in the application
domain---``create a new calendar event'' rather than ``click the button
that contains the text `new event'.''---which makes customizations
easier to build and more robust. However, the plugin API limits the
types of customizations that can be built, and many applications don't
have an extension system at all.

>>>>

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{End-User Software Customization by Direct Manipulation of Tabular
Data}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  In this paper we introduce \emph{table-driven customization}, a new
  way for end users to extend software by direct manipulation without
  doing traditional programming. We augment a user interface with a
  table view showing the structured data inside the application. When
  users edit the table, their changes are reflected in the original UI.
  This simple model accommodates a spreadsheet formula language and
  custom data editing widgets, providing enough power to implement a
  variety of useful extensions.

@@

Many end user customization systems
\citep{cook2007, bolin2005, leshed2008, chasins2018} offer a scripting
model. They use various strategies to make programming more
approachable: friendly syntax, a visual programming environment, or
macro recording to bootstrap from concrete demonstrations. But all these
techniques build on the same fundamental foundation: an imperative
programming model, with statements, mutable variables, and loops.

We have known for decades about an alternative: \emph{direct
manipulation} \citep{shneiderman1983}, where ``visibility of the object
of interest'' replaces ``complex command language syntax.'' Direct
manipulation is the \emph{de facto} standard in GUIs today, but when it
comes to customizing those GUIs, it is rarely to be found. Switching
from using an application to customizing it via scripting requires an
abrupt shift in interaction model, and can pose a steep learning barrier
for users not familiar with programming.

We subscribe to MacLean et al's vision of a ``gentle slope''
\citep{maclean1990} free of such abrupt ``cliffs,'' where users should
only need to make minimal and incremental investments in skill to
achieve their desired customizations. We seek to contribute to this
gentle slope with a new method for customizing software via direct
manipulation, which requires less skill than scripting but still
supports powerful customizations. We take inspiration from visual
database query interfaces and spreadsheets, which have successfully
enabled millions of end users to compute with data through direct
manipulation.

In our proposed paradigm, \emph{table-driven customization}, an
application's UI is augmented with a table view where the user can see
and manipulate the application's internal data. These changes don't just
apply to the table; they also result in immediate changes to the
application's original user interface. The user can sort/filter data in
the UI, inject annotations, pull in related information from other web
services, and more, all using the table as a mediating interface.
Interacting with the table view feels like interacting with a familiar
spreadsheet, but results in customizing an existing application.

@@

We have successfully used Wildcard to build customizations for 11
different websites that serve our own personal needs. In
Section~\ref{sec:reflections}, we present reflections from this process.
We outline the kinds of customizations we were able to build,
limitations we encountered, and some of the challenges of writing
scraping logic.

@@

\begin{itemize}
\tightlist
\item
  \emph{Customization by direct manipulation}: We discuss further how
  table-driven customization relates to a gentle slope of customization.
  We suggest that an important point on this slope is the ability to
  customize an application by directly seeing and changing its data,
  rather than by writing imperative scripts.
\item
  \emph{Semantic wrappers}: Typically, tools that don't rely on official
  extension APIs resort to offering low-level APIs for customization.
  Instead, we propose a community-maintained library of semantic
  wrappers around existing applications, enabling end users to work with
  domain objects rather than low-level representations.
\end{itemize}

Table-driven customization relates to existing work in many areas. Our
goals overlap with many software customization tools, and our methods
overlap with direct manipulation interfaces for working with structured
data, including visual database query systems and spreadsheets. We
explore these connections in Section~\ref{sec:related-work}.

Finanlly, in Section~\ref{sec:conclusion}, we conclude and describe some
opportunities for future work.

@@

To illustrate the end user experience of table-driven customization, we
consider an example scenario of customizing
\href{https://news.ycombinator.com/}{Hacker News}, a popular tech news
aggregator. Figure~\ref{fig:hacker-news} shows accompanying screenshots.

@@

To achieve this ordering, the user simply clicks on the ``points''
column header in the table. This sorts the table view by points, and the
website UI also becomes sorted in the same order
(Figure~\ref{fig:hacker-news}, Note B). Internally, Wildcard has changed
the website's DOM to synchronize it with the sort order of the table.
This sort predicate is also persisted in the browser and reapplied
automatically the next time the user loads the page, so they can always
browse the page sorted by points.

\textbf{Adding estimated read times}: Next, the user decides to attempt
a more substantial customization: to add estimated read times to each
article, in order to prioritize reading deeper content.

The table contains additional empty columns where the user can enter
spreadsheet-like formulas to compute derived values. The user enters a
formula into the first column, which is named \texttt{user1} by default
(Figure~\ref{fig:hacker-news}, Note C):

@@

This formula calls a built-in function that takes a URL as an argument
and uses a public web API to compute an estimated read time for the
URL's contents. The \texttt{link} argument in the formula refers to a
column name in the table; the formula is automatically evaluated across
all rows in the table, using the value of \texttt{link} for each row.

The user clicks the \texttt{user1} column header to sort the articles on
the page in descending order of estimated read time. They would also
like to display the read times in the page, but a number in seconds
isn't the most legible format, so they enter another formula in the
\texttt{user2} column:

@@

This formula converts seconds to minutes by dividing by 60 and rounding
to the nearest integer, and concatenates the result with a string label,
producing results like ``21 min read''.

Finally, the user clicks a menu option in the table header to display
the contents of this new column in the original page
(Figure~\ref{fig:hacker-news}, Note D). Each article on the page now
shows an annotation with the estimated read time in minutes. (The
formatting of annotations was determined by the programmer who created
the adapter for Hacker News.)

\textbf{Adding manual annotations}: The user can also manually add notes
to the table, by simply entering values into the table without using
formulas. In this case, the user jots down a few notes in another column
about articles they might want to read, and the notes appear in the page
next to the read times (Figure~\ref{fig:hacker-news}, Note D). The
annotations are also stored in the browser's local storage so they can
be retrieved on future visits.

@@

They can then filter the table to only contain rows where this formula
column contains \texttt{false}; links that the user has already visited
are hidden both from the table view and the original page. This is an
example of a customization that the original website could not have
implemented, since websites don't have access to the browser history for
privacy reasons. But by using Wildcard, the user could implement the
customization locally, without needing to expose their browser history
to Hacker News.

@@

Figure~\ref{fig:table-adapter} summarizes the overall architecture of
table-driven customization, using a simplified illustration of the
Hacker News example scenario. The name and points value for each article
is scraped from the web page DOM, and user annotations are loaded from
the brower's local storage.

First, the web page and the browser storage are each wrapped by a
\textbf{table adapter}, which defines a bidirectional mapping between an
underlying data source and a table.

In addition to a \emph{read mapping} for how the underlying data should
be represented as a table, it also has a \emph{write mapping} defining
the effects that edits have on the original data source. The local
storage adapter has a trivial mapping: it loads a table of data stored
in the browser, and persists edits to that state. The mapping logic of
the DOM scraping adapter is much more involved. It implements web
scraping logic to produce a table of data from the web page, and turns
edits into DOM manipulations, such as reordering rows of data on the
page.

@@

@@

\emph{Handling edits}: The query engine can issue a request to a table
adapter to make an edit to a record. The meaning of making an edit can
vary depending on the adapter: in the local storage adapter, a new value
may be persisted into local storage; in the DOM scraping adapter, an
edit may result in changing the value of a form field.

@@

\emph{Data from other tables}: The query engine provides each table
adapter with the entire combined table shown to the user. The DOM
scraping adapter uses this for injecting annotations---values from other
tables are added to the original web page.

\emph{Currently selected record}: As the user clicks around the table
view, the query engine broadcasts the record currently selected by the
user to each table adapter. The DOM scraping adapter uses this
information to highlight the row in the page that corresponds to the
selected row in the table, which helps the user understand the mapping
between the table and the original UI.

@@

Because every website has unique content, we rely on programmers to
create a DOM scraping adapter for each individual website to make it
available for customization in Wildcard. To make this approach viable,
we have built a generic DOM scraping adapter which a programmer can
configure with the minimal site-specific parts.

\emph{Basic metadata}: The programmer specifies URL patterns that the
adapter should activate on, and a set of column names and types to
return in the table.

\emph{Specifying scraping logic}: The main component of a site-specific
adapter is a single Javascript function that extracts relevant elements
from the page using web scraping technqiues. For each scraped row of
data, the programmer must return a unique identifier, and values for
each column in the table.

The generic adapter then wraps this scraping function to implement the
table adapter interface. For example, when the user sorts the table
view, the generic adapter synchronizes the sort order to the DOM by
removing the DOM elements corresponding to the rows in the table and
reinserting them back into the page in the new order.

\emph{Choosing an ID}: When possible, it is best to choose a server-side
identifier that remains stable across pageloads. This enables user
annotations persisted in local storage to be associated with the same
records on subsequent pageloads. We have found that it's usually
possible to find such an identifier; for example, each item in a page
often contains a link to a page with more details, with a URL that
contains a stable ID.

@@

The default option is to return a DOM element, in which case the generic
adapter extracts the text contents of the DOM element and casts them to
the type of the column. For example, if the scraping function returns a
\texttt{\textless{}div\textgreater{}} containing the string ``1.23'',
and the column has a numeric type, the generic adapter extracts the
number 1.23. The advantage of this approach is that the field can be
\emph{writable}---when the user edits the value in the table, the
generic adapter can simply overwrite the inner contents of the DOM
element to reflect the change. In practice, this is most useful for
editing the value of form input fields, although it's technically
possible to enable the editing of other fields (e.g., editing the title
of a post on Hacker News).

Another option is to directly return a value, rather than returning a
DOM element. The advantage of this approach is that the adapter author
can perform arbitrary computations to derive the returned value---for
example, they can use a regular expression to extract a substring. The
disadvantage is that the field is no longer writable. The computation
used to derive the value isn't reversible, so there's no way to reflect
a table edit in the DOM.

@@

As with DOM scraping adapters, we have made it easy for programmers to
create site-specific AJAX scraping adapters. A programmer writes a
function that specifies how to extract data from an AJAX request, and
the framework handles the details of intercepting requests and calling
the programmer-defined function.\footnote{So far we have only
  implemented AJAX scraping in the Firefox version of Wildcard, since
  Firefox has convenient APIs for intercepting requests. It appears
  possible to implement in Chrome and Edge as well, but we have not
  finished our implementation.}

@@

With the advent of rich frontend web frameworks, structured application
state is now often available in the web client. We suspect it is
possible to create plugins for frontend frameworks that expose this
state to Wildcard with only minimal effort from the application
developers. To test this hypothesis, we have created an early
proof-of-concept of an integrated website adapter for the popular
\href{https://redux.js.org/}{Redux} Javascript state management library.
To set up this integrated adapter, the programmer would only need to
make a few minor changes to their application code:

\begin{itemize}
\tightlist
\item
  \emph{Exposing a table}: In Redux, app developers already represent
  the state of a user interface as a single object. To expose a table of
  data to Wildcard, the developer simply writes a function that maps the
  state object to a table.
\item
  \emph{Editing data values}: The developer writes handlers which define
  how edits to table records should map to UI actions. For example,
  editing the value of a todo might trigger an \texttt{EDIT\_TODO}
  event.
\item
  \emph{Sorting/filtering the UI}: The developer inserts an additional
  data processing step before the state object is rendered in the UI, so
  that Wildcard can apply sorting and filtering.
\item
  \emph{Annotations}: The developer adds additional React components
  managed by the Wildcard adapter into their UI tree, which render
  additional data as annotations.
\end{itemize}

@@

One way to view this query model is as a tiny subset of the SQL query
model. Despite its simplicity, this model has proven sufficient for
meeting our customization needs, and minimizes the complexity of
supporting arbitrary queries. But because it fits into the general
paradigm of relational queries, it could theoretically be extended to
support a wider range of queries.

The query engine is also responsible for executing formulas. We have
built a small formula language resembling a spreadsheet formula
language, except that formulas automatically apply across an entire
column of data, and reference other column names instead of values in
specific rows. This is more convenient than needing to copy-paste a
formula across an entire column as in spreadsheets, and has worked for
all of the customizations we have built. Visual database query systems
like SIEUFERD \citep{bakke2016} use a similar approach to using formulas
in a structured table.

@@

The table editor only serves as a shallow interface layer over the query
engine, relaying user commands to the query engine and rendering the
resulting data table. Because of this architectural split, it would be
straightforward to develop additional table editor interfaces on top of
the Wildcard system. For example, we could provide a calendar view for
displaying a table containing a date column.

@@

We developed site-specific adapters for 11 websites that we personally
use frequently, and then built customizations in Wildcard for those
websites. Table 1 summarizes these results, showing the number of lines
of code in the adapter for each site, and some example customizations we
created. Here we offer our reflections from these experiences using the
system, focused on two key questions:

@@

Sometimes, websites have opaque ranking algorithms which presumably
maximize profit but restrict user agency. For example, Airbnb previously
allowed users to sort listings by price, but removed that feature in
2012. In other cases, a lack of sorting options seems more like an
innocent omission; for example, the Instacart grocery delivery service
has a spartan UI for viewing an order, which doesn't allow for sorting
items by price or category. In both of these cases, Wildcard enables
users to take back some control.

In the current implementation of Wildcard, users can only sort and
filter entries that are shown on the current page, which means that
users are not entirely liberated from the site's original ranking. This
restriction could be overcome in the future by scraping content across
multiple pages, or by using an integrated adapter and avoiding scraping
altogether. However, we've also realized that sorting/filtering a single
page of a paginated list is sometimes an acceptable outcome (and even a
preferable one). It's more useful, for example, to sort 30 recommended
Youtube videos than to try to sort all videos on Youtube.

\begin{figure}
\hypertarget{fig:amazon}{%
\centering
\includegraphics[width=\columnwidth]{media/amazon.png}
\caption{Sorting the used sellers page on Amazon by total price, including fees. The original page doesn't have sorting, and doesn't show the combined price.}\label{fig:amazon}
}
\end{figure}

\begin{figure}
\hypertarget{fig:ubereats}{%
\centering
\includegraphics[width=\columnwidth]{media/ubereats.png}
\caption{Organizing takeout restaurants on Uber Eats by delivery ETA and price}\label{fig:ubereats}
}
\end{figure}

@@

Many web annotation systems focus on annotating text or arbitrary
webpage content, but Wildcard limits annotations to structured objects
extracted by an adapter, resulting in a different set of use cases.
Annotating with Wildcard has proven most useful when taking notes on a
list of possible options (e.g., evaluating possible Airbnb locations to
rent). We have also used it with Instacart's online grocery cart, for
jotting down notes as we review an order and consider modifications
(shown in Figure~\ref{fig:instacart}).

@@

We developed two cell editors, to explore the benefits of adding custom
UI for editing values in the table.

One valuable feature is that users can incorporate their private
information into a web UI. We created a datepicker based on the
\href{https://fullcalendar.io/}{FullCalendar} plugin, which can load
data from a Google Calendar. This makes it convenient to enter dates
into a website based on the user's personal calendar information,
without uploading that information to the website itself.

Another benefit is that a user can choose a preferred widget for editing
information across different sites that typically provide a fragmented
editing experience. We built a text editor based on the
\href{https://ckeditor.com/}{CKEditor} rich text editor, and integrated
it with Google's Blogger website by representing the contents of a blog
post as a single table cell containing an HTML string (shown in
Figure~\ref{fig:blogger}). If there were site adapters available for
other websites with rich text entry, the user could then use a single
rich text editor to edit text across the whole web.

\begin{figure}
\hypertarget{fig:blogger}{%
\centering
\includegraphics[width=\columnwidth]{media/blogger.png}
\caption{Using a custom text editor widget to edit a blog post on Blogger. The text is synchronized with the Blogger editor through a table cell.}\label{fig:blogger}
}
\end{figure}

@@

AJAX scraping proved highly useful in several cases. The Uber Eats
website was challenging to scrape because it has a complex DOM structure
with machine-generated CSS classes, but the site also uses AJAX requests
which contain all the relevant data in a structured form that is much
easier to extract. We also found examples where relevant information
wasn't present in the DOM at all. On the grocery delivery site
Instacart, we found that AJAX requests contained additional data not
shown in the UI, enabling us to do things like sort grocery items by
category.

@@

Here we discuss some themes that we've explored with this work, which go
beyond our specific approach and address broader issues in software
customization.

@@

With table-driven customization we aim to provide a gentler slope, by
using direct manipulation for software customization. The data shown in
the table view is the domain data from the original application. The
user makes changes to the data by selecting areas of interest in the
table, e.g.~sorting/filtering by clicking the relevant column header, or
adding annotations by clicking and typing on the relevant row. At every
step, the user receives intermediate feedback, not only in the table
view, but also in the original application, so it's clear whether they
are making progress towards their desired result. These types of
interactions are common in GUI applications, and Wildcard therefore
seems to meet MacLean et al's goal: some one-click customizations are as
easy as using the original application. Formulas introduce some
additional complexity, but spreadsheets have demonstrated that formula
programming is still accessible to many users, due to the pure
functional semantics and the intermediate visibility of results.

One aspect of directness that we have chosen not to pursue in Wildcard
is enabling customization in closer proximity to the original user
interface elements, as explored by tools like Scotty \citep{eagan2011}.
We have found that augmenting the original UI with a distinct,
additional representation provides a more consistent experience across
all applications, and clearly shows the user what structured data is
available to work with.

Ainsworth et al provide a helpful taxonomy of the value of multiple
representations \citep{ainsworth1999}. In their terms, Wildcard plays a
\emph{complementary role} by supporting a \emph{different set of tasks}
from the original application, while displaying \emph{shared
information}. Wildcard may also help construct \emph{deeper
understanding by subtraction}---by stripping away details and only
showing the essential data in an interface, Wildcard encourages thinking
of an application in terms of its core information, rather than the
specific capabilities provided by the current user interface. In our
anecdotal experience, we've often found that looking at a site's data in
table format tends to spur new ideas for customizations which weren't
evident from looking at the original UI.

As noted in Section~\ref{sec:reflections}, there are many customizations
that can be achieved in scripting languages that cannot be implemented
in Wildcard. We consider this an acceptable tradeoff in exchange for a
gentler slope in customization, and we show in Section~\ref{sec:example}
and Section~\ref{sec:reflections} that our model can implement many
useful customizations.

@@

\emph{Anticipated customization tools}, in contrast, use explicit
extension APIs provided by the application developer. Examples of this
include accessing a backend web API, or writing a customization in
Applescript for an application that exposes its domain model to the
scripting language. The main benefit of this style is that it allows the
extension author to work with meaningful concepts in the application
domain---``create a new calendar event'' rather than ``click the button
that contains the text `new event'.''---which makes customizations
easier to build and more robust. However, the plugin API limits the
types of customizations that can be built, and many applications don't
have any plugin API.
